---
title: "Analysis"
author: "Nathan Green"
date: "17/03/2020"
output:
  pdf_document: 
    keep_tex: yes
    number_sections: yes
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The raw annual risk of death is calculated by the proportion of people in each age, sex, ethnicity group who die each year (with those who emigrate being censored, of course).
Separate survivorship curve for each year of birth, as life expectancy changes over time.
There is sex differences; ethnicity doesn't make much of a difference.

ETHPOP is based on ONS data in what year?


```{r comment=FALSE, warning=FALSE, message=FALSE}
library(readr)
library(purrr)
library(dplyr)
library(ggplot2)
library(scales)
library(reshape2)
library(survivorETHPOP)
```

# ETHPOP

We want to compare between the ONS mortality statistics and ETHPOP.
From [here](https://www.ons.gov.uk/peoplepopulationandcommunity/healthandsocialcare/healthandlifeexpectancies/methodologies/guidetocalculatingnationallifetables)
it details their method.
They calculate a _central rate of mortality_ as the average across 3 years.
$$
m_x = \sum_{y1,y2,y3} deaths_i/ \sum_{y1,y2,y3} pop_i
$$

Finally, they calculate the _mortality rate_ which is what we will be using to compare and is equivalent to the hazard.
$$
q_x = 2 m_x/(2 + m_x)
$$


## Individual categories hazards and survival

```{r comment=FALSE, warning=FALSE, message=FALSE}

ETHPOP_lifetable <- make_ETHPOP_lifetable()
# save(ETHPOP_lifetable, file = here::here("data", "ETHPOP_lifetable.RData"))


head(ETHPOP_lifetable)

ETHPOP_lifetable %>% 
  survivor_curve(group = list(sex = "M",
                              ETH.group = "WHO",
                              year = 2011)) %>% 
  haz_plot()

ETHPOP_lifetable %>% 
  survivor_curve(group = list(sex = "M",
                              ETH.group = "WHO",
                              year = 2049)) %>% 
  haz_plot(add = TRUE)

ETHPOP_lifetable %>% 
  survivor_curve(group = list(sex = "M",
                              ETH.group = "WHO",
                              year = 2011)) %>%
  haz_plot()

for (i in 2012:2043) {
  ETHPOP_lifetable %>% 
    survivor_curve(group = list(sex = "M",
                                ETH.group = "WHO",
                                year = i)) %>% 
    haz_plot(add = TRUE)
}

par(mfrow=c(1,2))
dat <- ETHPOP_lifetable %>% 
  survivor_curve(group = list(sex = "M",
                              ETH.group = "BAN",
                              year = 2011)) %>% 
  haz_plot() %>% 
  surv_plot()
```

### Ethnic groups

```{r comment=FALSE, warning=FALSE, message=FALSE}
ethnic_grps <- c("BAN", "BLA", "BLC", "CHI", "IND", "MIX",
                 "OAS", "OBL", "OTH", "PAK", "WBI", "WHO")

ETHPOP_lifetable %>% 
  survivor_curve(group = list(sex = "M",
                              ETH.group = "BAN",
                              year = 2011)) %>% 
  haz_plot()

for (i in ethnic_grps) {
  ETHPOP_lifetable %>% 
    survivor_curve(group = list(sex = "M",
                                ETH.group = i,
                                year = 2011)) %>% 
    haz_plot(add = TRUE)
}
```


```{r comment=FALSE, warning=FALSE, message=FALSE}

baseyr_2011 <- ETHPOP_lifetable$id[ETHPOP_lifetable$yr_age == "2011_0"][1]

ETHPOP_lifetable_2011M <- ETHPOP_lifetable %>% 
  filter(id == baseyr_2011,
         sex == "M")

ETHPOP_lifetable_2011F <- ETHPOP_lifetable %>% 
  filter(id == baseyr_2011,
         sex == "F")
```

```{r comment=FALSE, warning=FALSE, message=FALSE}
## ggplot

ggplot(ETHPOP_lifetable_2011M, aes(x = age, y = death_rate, colour = ETH.group)) +
  geom_line() +
  # scale_y_continuous(trans='log2') +
  coord_trans(y = "log10") +
  theme_bw()

ggplot(ETHPOP_lifetable_2011M, aes(x = age, y = qx, colour = ETH.group)) +
  geom_line() +
  # scale_y_continuous(trans='log2') +
  coord_trans(y = "log10") +
  theme_bw()

ggplot(ETHPOP_lifetable_2011M, aes(x = age, y = qx, colour = ETH.group)) +
  geom_line() +
  # scale_y_continuous(trans='log2') +
  theme_bw()

ggplot(ETHPOP_lifetable_2011M, aes(x = age, y = S, colour = ETH.group)) +
  geom_line() +
  theme_bw()
```

---

# ONS lifetables

Read in and check.

```{r comment=FALSE, warning=FALSE, message=FALSE}

lifetables <- read_lifetables()
# save(lifetables, file = here::here("data", "lifetables.RData"))


ONS_lifetables <-
  do.call(rbind, lifetables) %>% 
  mutate(new_yr = year < dplyr::lag(year, default = Inf),
         id = cumsum(new_yr)) %>% 
  group_by(id) %>% 
  mutate(baseyr = min(year)) %>% 
  ungroup() %>% 
  select(-id, -new_yr) %>% 
  mutate(baseyr = as.factor(baseyr))
```

```{r comment=FALSE, warning=FALSE, message=FALSE}

par(mfrow= c(2,3))
for (i in seq_along(lifetables)) {
  plot(x = lifetables[[i]]$year[lifetables[[i]]$sex == "M"],
       y = lifetables[[i]]$qx[lifetables[[i]]$sex == "M"], log = "y", type = "l",
       main = names(lifetables)[i], ylab = "h", xlab = "year")
  lines(x = lifetables[[i]]$year[lifetables[[i]]$sex == "F"],
        y = lifetables[[i]]$qx[lifetables[[i]]$sex == "F"], log = "y", type = "l", col = "red")
}

par(mfrow= c(1,2))
for (j in c("M","F")) {
  plot(x = lifetables[[1]]$age[lifetables[[1]]$sex == j],
       y = lifetables[[1]]$qx[lifetables[[1]]$sex == j], log = "y", type = "l",
       main = j, ylab = "h", xlab = "age")
  for (i in seq_along(lifetables)) {
    lines(x = lifetables[[i]]$age[lifetables[[i]]$sex == j],
          y = lifetables[[i]]$qx[lifetables[[i]]$sex == j], log = "y", type = "l", col = i)
  }
}
```

```{r comment=FALSE, warning=FALSE, message=FALSE}

ggplot(ONS_lifetables, aes(x = age, y = qx, colour = interaction(sex, baseyr))) +
  geom_line() +
  # scale_y_continuous(trans='log2') +
  coord_trans(y = "log10") +
  theme_bw()

```

# Comparison with ONS and ETHPOP

2011

```{r comment=FALSE, warning=FALSE, message=FALSE}

ggplot(ETHPOP_lifetable_2011M, aes(x = age, y = death_rate, colour = ETH.group)) +
  geom_line() +
  coord_trans(y = "log10") +
  ggtitle("Male 2011") +
  theme_bw() +
  geom_line(aes(age, qx, colour = "ONS"),
            data = ONS_lifetables[ONS_lifetables$baseyr == 2011 & ONS_lifetables$sex == "M", ],
            colour = "black")

ggplot(ETHPOP_lifetable_2011F, aes(x = age, y = death_rate, colour = ETH.group)) +
  geom_line() +
  ggtitle("Female 2011") +
  coord_trans(y = "log10") +
  theme_bw() +
  geom_line(aes(age, qx, colour = "ONS"),
            data = ONS_lifetables[ONS_lifetables$baseyr == 2011 & ONS_lifetables$sex == "F", ],
            colour = "black")
```

```{r comment=FALSE, warning=FALSE, message=FALSE}

ggplot(ETHPOP_lifetable_2011M, aes(x = age, y = death_rate, colour = ETH.group)) +
  geom_line() +
  ggtitle("Male 2011") +
  theme_bw() +
  geom_line(aes(age, qx, colour = "ONS"),
            data = ONS_lifetables[ONS_lifetables$baseyr == 2011 & ONS_lifetables$sex == "M", ],
            colour = "black") +
  ylim(0, 0.003) + xlim(0, 60)

ggplot(ETHPOP_lifetable_2011F, aes(x = age, y = death_rate, colour = ETH.group)) +
  geom_line() +
  ggtitle("Female 2011") +
  theme_bw() +
  geom_line(aes(age, qx, colour = "ONS"),
            data = ONS_lifetables[ONS_lifetables$baseyr == 2011 & ONS_lifetables$sex == "F", ],
            colour = "black") +
  ylim(0, 0.003) + xlim(0, 60)
```

2030

```{r comment=FALSE, warning=FALSE, message=FALSE}

baseyr_2030 <- ETHPOP_lifetable$id[ETHPOP_lifetable$yr_age == "2030_0"][1]

ETHPOP_lifetable[ETHPOP_lifetable$id == baseyr_2030 & ETHPOP_lifetable$sex == "M", ] %>% 
  ggplot(aes(x = age, y = death_rate, colour = ETH.group)) +
  geom_line() +
  coord_trans(y = "log10") +
  ggtitle("Male 2030") +
  theme_bw() +
  geom_line(aes(age, qx, colour = "ONS"),
            data = ONS_lifetables[ONS_lifetables$baseyr == 2030 & ONS_lifetables$sex == "M", ],
            colour = "black")

ETHPOP_lifetable[ETHPOP_lifetable$id == baseyr_2030 & ETHPOP_lifetable$sex == "F", ] %>% 
  ggplot(aes(x = age, y = death_rate, colour = ETH.group)) +
  geom_line() +
  ggtitle("Female 2030") +
  coord_trans(y = "log10") +
  theme_bw() +
  geom_line(aes(age, qx, colour = "ONS"),
            data = ONS_lifetables[ONS_lifetables$baseyr == 2030 & ONS_lifetables$sex == "F", ],
            colour = "black")
```

```{r comment=FALSE, warning=FALSE, message=FALSE}

baseyr_2030 <- ETHPOP_lifetable$id[ETHPOP_lifetable$yr_age == "2030_0"][1]

ETHPOP_lifetable[ETHPOP_lifetable$id == baseyr_2030 & ETHPOP_lifetable$sex == "M", ] %>% 
  ggplot(aes(x = age, y = death_rate, colour = ETH.group)) +
  geom_line() +
  ggtitle("Male 2030") +
  theme_bw() +
  geom_line(aes(age, qx, colour = "ONS"),
            data = ONS_lifetables[ONS_lifetables$baseyr == 2030 & ONS_lifetables$sex == "M", ],
            colour = "black") +
  ylim(0, 0.003) + xlim(0, 40)

ETHPOP_lifetable[ETHPOP_lifetable$id == baseyr_2030 & ETHPOP_lifetable$sex == "F", ] %>% 
  ggplot(aes(x = age, y = death_rate, colour = ETH.group)) +
  geom_line() +
  ggtitle("Female 2030") +
  theme_bw() +
  geom_line(aes(age, qx, colour = "ONS"),
            data = ONS_lifetables[ONS_lifetables$baseyr == 2030 & ONS_lifetables$sex == "F", ],
            colour = "black") +
  ylim(0, 0.003) + xlim(0, 40)
```


## Life expectancy

ONS

```{r}

ggplot(ONS_lifetables[ONS_lifetables$sex == "M", ], aes(age, ex, colour = baseyr)) +
  geom_line() +
  theme_bw()

ggplot(ONS_lifetables[ONS_lifetables$sex == "F", ], aes(age, ex, colour = baseyr)) +
  geom_line() +
  theme_bw()
```


```{r eval=FALSE}
rmarkdown::render("vignettes/analysis.Rmd", output_dir = "docs/", output_format = "html_document")
```
